---
title: "Bike Share TO Ride Report"
output: html_document
---

```{r header, echo=FALSE, warning=FALSE, include=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

source("R/mapper_funcs.R")
# Load bike share stations ####
stations <- get_bike_share_data()

# Load data ####
input_file <- "sample_data/ride_data.txt"

data <- get_ride_data(input_file)
```

*`r format(Sys.time(), "%B %d, %Y")`*

Ride data: ``r input_file``

## Summary Statistics

```{r stats, include=FALSE, cache=TRUE}
freq <- calculate_station_frequencies(data, stations)

most_start <- which(freq$N == max(freq[freq$type=="Start", "N"]) & 
                      freq$type == "Start")
most_end <- which(freq$N == max(freq[freq$type=="End", "N"]) & 
                    freq$type == "End")

avg_ride <- round(mean(data$Duration), digits=2)
shortest_trip<- which.min(data$Duration)
longest_trip <- which.max(data$Duration)
most_freq_trips <- data %>% 
  group_by(Start.Station, End.Station) %>% 
  summarize(N = length(Start.Station), 
            Avg.Duration = round(mean(Duration), digits=2)) %>% 
  mutate(route = paste(Start.Station, "to", End.Station)) %>%
  as.data.frame %>%
  arrange(desc(N))

ddf <- format_by_datetime(data)

trips_in_day <- ddf %>% group_by(dy, mo, yr) %>% 
  summarize(N = length(route))
  
most_trips_in_day <- trips_in_day %>%
  subset(N == max(N)) %>%
  mutate(date = paste(mo, dy, yr, sep="/")) %>% 
  as.data.frame
```

- Total rides: `r nrow(data)`
- Most frequent start station, visits: `r freq[most_start, c("stationName", "N")]`
- Most frequent end station, visits: `r freq[most_end, c("stationName", "N")]`
- Average ride time: `r avg_ride` minutes
- Shortest trip: `r data[shortest_trip, "Start.Station"]` to 
`r data[shortest_trip, "End.Station"]`, `r data[shortest_trip, "Duration"]` minutes
- Longest trip: `r data[longest_trip, "Start.Station"]` to 
`r data[longest_trip, "End.Station"]`, `r data[longest_trip, "Duration"]` minutes
- Average number of rides per day: `r mean(trips_in_day$N)`
- Most trips in one day: `r unique(most_trips_in_day$N)` (`r most_trips_in_day[,"date"]`)
- Most popular routes:
    1. `r most_freq_trips[1, "route"]`, `r most_freq_trips[1, "N"]`
    1. `r most_freq_trips[2, "route"]`, `r most_freq_trips[2, "N"]`
    1. `r most_freq_trips[3, "route"]`, `r most_freq_trips[3, "N"]`

```{r echo=FALSE, warning=FALSE, fig.width=8, fig.height=8, cache=TRUE}
i <- 4
print(rbind(inner_join(
  data.frame(stationName = most_freq_trips[1:i, "Start.Station"],
             most_freq_trips[1:i, c("route", "N")]), 
  freq[,c("stationName", "longitude", "latitude")]),
  inner_join(
    data.frame(stationName = most_freq_trips[1:i, "End.Station"],
               most_freq_trips[1:i, c("route", "N")]), 
    freq[,c("stationName", "longitude", "latitude")])) %>%
  unique %>%
  map_stations + facet_wrap(~ route, ncol = 2) + ggtitle("Most popular routes"))
```



```{r circos, fig.width=10, fig.height=6, eval=FALSE}
library(circlize)
library(RColorBrewer)
factors <- c(most_freq_trips[1:5, "Start.Station"], 
             most_freq_trips[1:5, "End.Station"]) %>%  unique
factors <- factor(factors, levels=factors)
circos.initialize(factors, xlim = c(0,10))
circos.trackPlotRegion(factors = factors, 
                       ylim = c(0, 1), 
                       bg.col = brewer.pal(length(factors), "Set1"))

for (i in 1:length(factors)) {
  circos.text(5, 1, factors[1], sector.index=factors[i], track.index=1,
              niceFacing = TRUE)
}

ribbon.col <- brewer.pal(5, "Pastel1")
for (i in 1:5) {
  circos.link(most_freq_trips[i, "Start.Station"], c(1,3), 
              most_freq_trips[i, "End.Station"], c(7,9),
              col = ribbon.col[i],
              lwd = most_freq_trips[i, "N"])
}

```


## Map
```{r ride_map, fig.width=10, fig.height=6, cache=TRUE}
print(map_stations(df = freq) + facet_wrap(~ type, ncol = 2) +
  ggtitle("Map of visited stations"))
```

```{r ride_map_merged, fig.width=7, fig.height=6, cache=TRUE}
print(map_stations(df = aggregate(N ~ stationName + latitude + longitude, 
                                  data = freq, sum)) + 
        ggtitle("Map of all visited stations"))
```

## Graphs
```{r plot_by_month, fig.width=8, fig.height=5}
ddf %>% group_by(yr, mo, mo.abbr) %>% 
  summarize(N = length(mo)) %>%
  arrange(yr, mo) %>%
  ggplot(aes(x = mo.abbr, 
             y = N, 
             fill = yr)) + 
  geom_bar(stat = "identity", position="dodge") +
  xlab("Month") + ylab("Number of trips") +
  ggtitle("Number of trips by month and year")
```

```{r plot_trip_by_day, fig.width=10, fig.height=7, cache=TRUE}
ddf %>% group_by(yr, mo, mo.abbr, dy) %>%
  summarize(avg.duration = round(mean(duration), digits=2),
            sd = round(sd(duration), digits=2)) %>%
  ggplot(aes(x = dy,
             y = avg.duration, color=yr, group=yr)) +
  geom_point() +
  geom_line() +
  geom_errorbar(aes(ymin = avg.duration - sd, ymax = avg.duration + sd, 
                    color = yr), 
                width = 0.2, position=position_dodge(width = 0.1)) +
  facet_wrap(~ mo.abbr, ncol = 3, scale="free_x") +
  theme(axis.text.x = element_text(angle= 45, hjust = 1, size=8)) +
  xlab("Day") + ylab("Duration") +
  ggtitle("Trip duration by day")
```

```{r plot_trip_by_month, fig.width=10, fig.height=5}
ddf %>%
  ggplot(aes(x = mo.abbr,
             y = duration, fill=yr)) +
  geom_boxplot(position="dodge") +
  xlab("Month") + ylab("Duration") +
  ggtitle("Trip duration by month")
```

```{r plot_trip_by_station, fig.width=10, fig.height=10}
ddf %>% group_by(yr, mo, mo.abbr, route) %>%
  summarize(N = length(route),
            avg.duration = round(mean(duration), digits=2),
            sd = round(sd(duration), digits=2)) %>%
  subset(route %in% most_freq_trips[1:6, "route"]) %>%
  ggplot(aes(x = mo.abbr,
             y = avg.duration, group=yr)) +
  geom_point(aes(color=yr)) +
  geom_line(aes(color=yr)) +
  geom_errorbar(aes(ymin = avg.duration - sd, ymax = avg.duration + sd, 
                    color = yr), 
                width = 0.2, position=position_dodge(width = 0.1)) +
  facet_wrap(~ route, ncol =2, scale="free_x") +
  xlab("Month") + ylab("Average Duration") +
  ggtitle("Trip duration by most frequent routes")
```

```{r plot_time_of_day, fig.width=10, fig.height=5}
ddf %>% 
  mutate(HR=as.numeric(as.character(hr)) + 
           as.numeric(as.character(min))/60) %>%
  ggplot(aes(x = HR)) +
  geom_histogram(binwidth = 0.25) +
  xlab("Time (24 hr)") + ylab("Count") +
  ggtitle("Trips by time of day") +
  scale_x_continuous(breaks = seq(1,24,1))
```

```{r plot_num_trips_by_day, fig.width=10, fig.height=5}
ggplot(trips_in_day, aes(x = N, fill = yr, group = yr)) +
  geom_histogram(position="dodge", binwidth = 1) +
  xlab("Trips per day") + ylab("Count") +
  ggtitle("Number of trips per day")
```

<center>
<hr>
&copy; 2014, 2015 Kevin Ha

https://github.com/kcha/bike_share_ride_report
</center>
